---
title: "Lab Exercises Week 03"
author: "Alison Lawyer"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include = TRUE}
### keep this chunk in all your RMarkdown scripts
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 60), tidy = TRUE)
```


```{r}
# List required packages
library(tidyverse)
library(ggplot2)
library(palmerpenguins)
```



For this lab, you will create various data visualizations of the palmerpenguins dataset to practice your understanding of ggplot and the grammar of graphics and apply visualization best practices.

```{r}
# load the palmerpenguins data
# this time, you can use the cleaned dataset called 'penguins'
data("penguins")


```



# Part 1

## Exercise 1

You have already created histograms previously using the hist() function from base R. Now you will learn how to do so with ggplot()

```{r}
# Create a histogram of flipper length using geom_histogram()
# Remember to start with the basic ggplot layer and add aesthetics first. What do you put inside aes()? 
ggplot(na.omit(penguins), aes(x = flipper_length_mm)) + 
  geom_histogram()
## OR
ggplot(penguins, aes(x = flipper_length_mm)) + 
  geom_histogram(na.rm = TRUE)

ggplot(na.omit(penguins), aes(x = flipper_length_mm)) + 
  geom_histogram(bins = 50, color = "black", fill = "lightgreen") +
    labs(title = "Histogram of flipper length", 
       x = "Flipper length (g)", 
       y = "Count") +
  theme_minimal()

```


## Exercise 2

In this exercise you will practice modifying the look of your graph using a variable from the data. 

```{r}
# Modify your histogram of flipper length by coloring the bars by species
# Remember that to achieve this you will want to add a variable signifying species into the aes() function of ggplot

ggplot(na.omit(penguins), aes(x = flipper_length_mm, fill = species)) + 
  geom_histogram(bins = 50) +
    labs(title = "Histogram of flipper length", 
       x = "Flipper length (g)",
       y = "Count") +
  theme_minimal()



# You will see some of the bars for different species overlap
# To deal with this, within the geom_histogram(), which draws your bars, use transparency (alpha) on the fill of the bars so that overlapping bars are visible
ggplot(na.omit(penguins), aes(x = flipper_length_mm, fill = species)) + 
  geom_histogram(position = "identity", alpha = 0.5, bins = 40, color = "black") +
    labs(title = "Histogram of flipper length", 
       x = "Flipper length (g)", 
       y = "Count") +
  theme_minimal()

```


## Exercise 3

Now experiment with the facet_wrap and facet_grid functions to create a histogram of flipper lengths for all three penguin species, but instead of one figure with overlapping bars you will now separate out the histogram for each species.  

```{r}
# Instead of coloring bars by species, it is cleaner to create separate graphs for each species
# Use the facet_wrap() function in ggplot to create multiples of a figure for a grouping variable. 
ggplot(na.omit(penguins), aes(x = flipper_length_mm, fill = species)) + 
  theme_minimal() +
  geom_histogram(bins = 50, color = "black") +
    labs(title = "Histogram of flipper length", 
       x = "Flipper length (g)", 
       y = "Count") +
  facet_wrap(~species, nrow = 3)


# Now explore the use of facet_grid instead and notice the difference in behavior of the resulting plots. 
ggplot(na.omit(penguins), aes(x = flipper_length_mm, fill = species)) + 
  theme_minimal() +
  geom_histogram(bins = 50, color = "black") +
    labs(title = "Histogram of flipper length", 
       x = "Flipper length (g)", 
       y = "Count") +
  facet_grid(cols = vars(species))



```



## Exercise 4

Now let's explore the sexual dimorphism among Adelie penguins. Create a graph of the mean body mass by sex.

```{r}
# Create a data summary by sex and species that calculates the mean and standard deviation of body mass
adelie_mass_summary <- na.omit(penguins) %>%
  filter(species == "Adelie") %>%
  group_by(sex) %>%
  summarize(mean_body_mass = mean(body_mass_g),
            sd_body_mass = sd(body_mass_g))

# Plot the mean body mass for males and females for adelie penguins.
# Add appropriate labels. 
adelie_mass_summary %>%
  ggplot(aes(x = sex, y = mean_body_mass)) +
  theme_minimal() +
  geom_col(fill = "darkblue") +
    labs(title = "Plot of mean body mass", 
       x = "Sex", 
       y = "Body mass (g)")
  



# Apply a theme of your choice to the graph. You can explore different themes by typing ?theme and going to complete themes or by adding a theme to your plot starting with theme_ and trying out the different options that pop up. There are even more themes available when you install and load the package 'ggthemes'

## already been using theme in functions above.


```


## Exercise 5

A common element of bar charts that plot mean values is to add a measure of variation. 
In this exercise you will practice adding errorbars to a graph using the standard deviation from your data summary. 

```{r}
# Add error bars to the graph. Error bars should represent one standard deviation. 
# Note: Use the geometry geom_errorbar() for creating your error bars
adelie_mass_summary %>%
  ggplot(aes(x = sex, y = mean_body_mass, fill = sex)) +
  theme_minimal() +
  geom_col() +
    labs(title = "Plot of mean body mass", 
       x = "Sex", 
       y = "Body mass (g)") +
  geom_errorbar(aes(
    ymin = mean_body_mass - sd_body_mass, 
    ymax = mean_body_mass + sd_body_mass), 
    width = 0.3)


```


