---
title: "Homework Exercises Week 08"
author: "Alison Lawyer"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include = TRUE}
# keep this chunk in all your RMarkdown scripts
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 20), tidy = TRUE)
```

```{r}
# List required packages
library(tidyverse)
library(openxlsx)
library(kableExtra)
library(ggplot2)

```



# HOMEWORK EXERCISES


## Exercise 1

Conduct exploratory analyses for the question posed in exercise 3 from the lab. The steps and outputs should mirror what you did for exercise 1, but adjusted for the question posed and the kind of analysis done. For example, think through what you are comparing and then decide how best to structure your table of descriptive stats, graphs, and tests of normality. Determine what to use for the visual assessment of normal distribution referring to the lecture and readings for this week. Write a summary statement about your observations, including whether assumptions of the t-test you conducted in exercise 3 were met. Include all pertinent information in your summary statement. 

```{r}
# Load the data
vigilance <- read.xlsx("vigilance.xlsx")

# Clean column names
colnames(vigilance) <- sub(pattern = "\\.\\((.*?)\\)", replacement = '', colnames(vigilance))

ct_bt <- vigilance %>% 
  filter(species == "CT" | species == "BT")

control <- vigilance %>%
  filter(treatment == "control",
         species == "CT" | species == "BT") %>%
  select(site, species, treatment, vigilance.rate) %>%
  pivot_wider(values_from = "vigilance.rate",
              names_from = c("species", "treatment")) %>%
  mutate(control_differences = BT_control - CT_control)

urban <- vigilance %>%
  filter(treatment == "urban noise",
         species == "CT" | species == "BT") %>%
  select(site, species, treatment, vigilance.rate) %>%
  pivot_wider(values_from = "vigilance.rate",
              names_from = c("species", "treatment")) %>%
  mutate(urban_differences = `BT_urban noise` - `CT_urban noise`)

combined <- as.data.frame(c(control %>% select(site, control_differences),
                            urban %>% select(urban_differences))) %>% 
  pivot_longer(cols = control_differences:urban_differences,
                     names_to = "treatment",
                     values_to = "diffs")


# Produce summary table. Be sure to group by species and treatment. 
summary_table <- ct_bt %>%
  group_by(species, treatment) %>%
  summarize(mean_vigilance = mean(vigilance.rate),
            sd_vigilance = sd(vigilance.rate),
            median_vigilance = median(vigilance.rate),
            iqr = IQR(vigilance.rate),
            var = var(vigilance.rate),
            se = sd_vigilance/sqrt(n()),
            CI_low = mean_vigilance - 1.96*se,
            CI_high = mean_vigilance + 1.96*se
  )
summary_table

# Make nice table for printing. Note that your primary goal is to compare species within each condition, which is different 
# from Exercise 1. Think about how to best arrange the summary table to make this comparison as easy as possible.
summarized <- summary_table %>% 
  pivot_longer(cols = "mean_vigilance":"CI_high",
               names_to = "stat",
               values_to = "values") %>%
  pivot_wider(names_from = species, values_from = values)

pretty_summary <- kable(summarized, caption = "Summary statistics by treatment and species")
# Print table
pretty_summary


# Figure comparing the means
summary_table %>% 
  ggplot(aes(x = species, y = mean_vigilance, fill = species)) +
  geom_col(show.legend = FALSE) +
  geom_errorbar(aes(ymin = CI_low, ymax = CI_high), width = 0.3, color = "black") +
  facet_wrap(~treatment) +
  labs(title = "Mean vigilance rates of BT and CT species by noise treatment condition",
       x = "Species",
       y = "Mean vigilance rate",
       caption = "error bars represent a 95% confidence interval")

# Histograms
# Here what matters is a histogram of the differences. You can add to the pivoted wide table created above. 
ggplot(combined, aes(x = diffs, fill = treatment)) +
  geom_histogram(binwidth = 3, position = "dodge", na.rm = TRUE, show.legend = FALSE) +
  labs(title = "Distribution of differences between BT and CT species vigilance rates by treatment",
       x = "Differences",
       y = "Count")

ggplot(combined, aes(x = diffs, fill = treatment)) +
  geom_density(alpha = 0.5, na.rm = TRUE, show.legend = FALSE) +
  labs(title = "Distribution of differences between BT and CT species vigilance rates by treatment",
       x = "Differences")


# Q-Q plots
# Be sure to do these for the differences as well
ggplot(na.omit(combined), aes(sample = diffs, color = treatment)) +
  stat_qq(show.legend = FALSE) +
  geom_abline(aes(intercept = mean(x = diffs), 
                  slope = sd(x = diffs))) +
    labs(title = "Q-Q plots of difference between BT and CT by noise treatment",
       x = "Theoretical quantiles",
       y = "Sample quantiles",
       caption = "y-intercept is mean differences, slope is standard deviation")


```

### E1 Summary:
Based on our graph comparing the means of each species for each noise condition, we can see that there is a larger difference between the Blue Tits and Coal Tits in the urban noise condition. Based on our histograms and Q-Q plots, we can see that the differences in the vigilance rates between the species is roughly normally distributed in each noise condition.
Based on our paired t-tests (run in the lab exercise 3), we found that the differences between the two species' mean vigilance rates are not statistically significant [t(24) = -1.35, p = 0.19] in the control condition. However in the urban noise condition, the two species exhibit statistically different vigilance rates [t(24) = -3.63, p <0.01]. Blue Tits were likely to have vigilance rates 3.04 to 11.05 lower than Coal Tits (95% CI), with a mean difference 7.04 lower.
We were able to run the paired t-tests because our data consists of continuous variables, our observations are independent while the groups are dependent, and the differences are roughly normally distributed.


## Exercise 2

Similar to exercise 1 above, compare the vigilance behavior between control and noise treatment sites for all other species present in the data. Here, simply check whether assumption for the t-test are met, then conduct the t-tests, and report on the results. No need to produce tables of descriptive statistics for each species. In which species is the difference the greatest and smallest? Be sure to report the difference in mean vigilance rates and their 95% confidence. 

```{r}
GT <- vigilance %>%
  filter(species == "GT") %>%
  select(site, species, treatment, vigilance.rate) 

ggplot(GT, aes(sample = vigilance.rate)) +
  stat_qq() +
  geom_abline(aes(intercept = mean(x = vigilance.rate), 
                  slope = sd(x = vigilance.rate))) +
  labs(title = "Q-Q plot of GT vigilance rate",
   x = "Theoretical quantiles",
   y = "Sample quantiles",
   caption = "y-intercept is mean vigilance rate, slope is standard deviation")

t.test(formula = GT$vigilance.rate ~ GT$treatment) 

RB <- vigilance %>%
  filter(species == "RB") %>%
  select(site, species, treatment, vigilance.rate) 

ggplot(RB, aes(sample = vigilance.rate)) +
  stat_qq() +
  geom_abline(aes(intercept = mean(x = vigilance.rate), 
                  slope = sd(x = vigilance.rate))) +
  labs(title = "Q-Q plot of RB vigilance rate",
   x = "Theoretical quantiles",
   y = "Sample quantiles",
   caption = "y-intercept is mean vigilance rate, slope is standard deviation")

t.test(formula = RB$vigilance.rate ~ RB$treatment) 

CH <- vigilance %>%
  filter(species == "CH") %>%
  select(site, species, treatment, vigilance.rate) 

ggplot(CH, aes(sample = vigilance.rate)) +
  stat_qq() +
  geom_abline(aes(intercept = mean(x = vigilance.rate), 
                  slope = sd(x = vigilance.rate))) +
  labs(title = "Q-Q plot of CH vigilance rate",
   x = "Theoretical quantiles",
   y = "Sample quantiles",
   caption = "y-intercept is mean vigilance rate, slope is standard deviation")

t.test(formula = CH$vigilance.rate ~ CH$treatment) 

NH <- vigilance %>%
  filter(species == "NH") %>%
  select(site, species, treatment, vigilance.rate) 

ggplot(NH, aes(sample = vigilance.rate)) +
  stat_qq() +
  geom_abline(aes(intercept = mean(x = vigilance.rate), 
                  slope = sd(x = vigilance.rate))) +
  labs(title = "Q-Q plot of NH vigilance rate",
   x = "Theoretical quantiles",
   y = "Sample quantiles",
   caption = "y-intercept is mean vigilance rate, slope is standard deviation")

t.test(formula = NH$vigilance.rate ~ NH$treatment) 


```

### E2 Summary:
We are able to run independent t-tests here because our data consists of continuous variables, our observations are independent along with each group being independent, and the rates are roughly normally distributed.
Based on our independent t-tests, we can conclude the following: 
- There is not a statistically significant difference between vigilance rates in the control and noise treatments for the GT bird species [t(71.77)) = -1.67, p = 0.09]. The mean vigilance rate is likely to vary by 10 to 13 (95% CI).
- There is not a statistically significant difference between vigilance rates in the control and noise treatments for European Robins [t(41.97) = -0.49, p = 0.63]. The mean vigilance rate is likely to vary by -9 to 5 (95% CI). This species has the smallest difference between the mean vigilance rates between treatment sites, compared to the other species. 
- There is not a statistically significant difference between vigilance rates in the control and noise treatments for the CH bird species [t(7.85) = -1.31, p = 0.22]. The mean vigilance rate is likely to vary by -25 to 6 (95% CI). This species has the greatest difference between the mean vigilance rates between treatment sites, compared to the other species.
- There is a statistically significant difference between vigilance rates in the control and noise treatments for the NH bird species [t(14.57) = -2.91, p = 0.01]. The mean vigilance rate is likely to vary by -10 to -1 (95% CI).

## Exercise 3

Assess whether there is a difference in wind speed, temperature and hours from sunrise between control and noise treatment sites. This could of course influence the outcome of the study by influencing vigilance behavior of any birds that are being observed. For each of these three continuous variables, conduct descriptive analyses and summary, a visual assessment of normal distribution, conduct an appropriate t-test, and report your results in a summary statement. 

```{r}
## WIND SPEED
wind_speed <- vigilance %>%
  select(site, species, treatment, wind.speed) 

wind_summary <- wind_speed %>%
  group_by(treatment) %>%
  summarize(mean_wind = mean(wind.speed),
            sd_wind = sd(wind.speed),
            median_wind = median(wind.speed),
            iqr = IQR(wind.speed),
            var = var(wind.speed),
            se = sd_wind/sqrt(n()),
            CI_low = mean_wind - 1.96*se,
            CI_high = mean_wind + 1.96*se
  )
wind_summary

summarized_wind <- wind_summary %>% 
  pivot_longer(cols = "mean_wind":"CI_high",
               names_to = "stat",
               values_to = "values") %>%
  pivot_wider(names_from = treatment, values_from = values)

pretty_summary_wind <- kable(summarized_wind, caption = "Summary statistics by treatment for wind speed")
pretty_summary_wind


# Figure comparing the means
wind_summary %>% 
  ggplot(aes(x = treatment, y = mean_wind, fill = treatment)) +
  geom_col(show.legend = FALSE) +
  geom_errorbar(aes(ymin = CI_low, ymax = CI_high), width = 0.3, color = "black") +
  labs(title = "Mean wind speed by noise treatment condition",
       x = "Treatment condition",
       y = "Mean wind speed",
       caption = "error bars represent a 95% confidence interval")

# Histograms
ggplot(wind_speed, aes(x = wind.speed, fill = treatment)) +
  geom_histogram(binwidth = 1, position = "dodge", na.rm = TRUE, show.legend = FALSE) +
  labs(title = "Distribution of wind speed by treatment",
       x = "Wind speed",
       y = "Count")

ggplot(wind_speed, aes(x = wind.speed, fill = treatment)) +
  geom_density(alpha = 0.5, na.rm = TRUE, show.legend = FALSE) +
  labs(title = "Distribution of wind speed by treatment",
       x = "Wind speed")

# Q-Q plots
ggplot(wind_speed, aes(sample = wind.speed, color = treatment)) +
  stat_qq(show.legend = FALSE) +
  geom_abline(aes(intercept = mean(x = wind.speed), 
                  slope = sd(x = wind.speed))) +
  labs(title = "Q-Q plot of wind speed by treatment",
   x = "Theoretical quantiles",
   y = "Sample quantiles",
   caption = "y-intercept is mean wind speed, slope is standard deviation")

t.test(formula = wind_speed$wind.speed ~ wind_speed$treatment)
```

### E3 Wind Speed Summary:
Based on our graph comparing the means of the wind speeds by treatment condition, we can see that there isn't a large difference between the means for each condition. Based on our histograms and Q-Q plots, we can see that both treatment conditions are heavily right-skewed.
Though the data are skewed, we ran independent t-tests here because our data consists of continuous variables and our observations are independent along with each group being independent.
Based on our independent t-tests, there is not a statistically significant difference between vigilance rates in the control and noise treatments for the GT bird species [t(235)) = -0.55, p = 0.05]. The mean vigilance rate is likely to vary by -1.1 to 0.6 (95% CI).


```{r}
## TEMPERATURE
temperature <- vigilance %>%
  select(site, species, treatment, temperature) 

temperature_summary <- temperature %>%
  group_by(treatment) %>%
  summarize(mean_temp = mean(temperature),
            sd_temp = sd(temperature),
            median_temp = median(temperature),
            iqr = IQR(temperature),
            var = var(temperature),
            se = sd_temp/sqrt(n()),
            CI_low = mean_temp - 1.96*se,
            CI_high = mean_temp + 1.96*se
  )
temperature_summary

summarized_temp <- temperature_summary %>% 
  pivot_longer(cols = "mean_temp":"CI_high",
               names_to = "stat",
               values_to = "values") %>%
  pivot_wider(names_from = treatment, values_from = values)

pretty_summary_temp <- kable(summarized_temp, caption = "Summary statistics by treatment for temperature")
pretty_summary_temp


# Figure comparing the means
temperature_summary %>% 
  ggplot(aes(x = treatment, y = mean_temp, fill = treatment)) +
  geom_col(show.legend = FALSE) +
  geom_errorbar(aes(ymin = CI_low, ymax = CI_high), width = 0.3, color = "black") +
  labs(title = "Mean temperature by noise treatment condition",
       x = "Treatment condition",
       y = "Mean temperature",
       caption = "error bars represent a 95% confidence interval")

# Histograms
ggplot(temperature, aes(x = temperature, fill = treatment)) +
  geom_histogram(binwidth = 1, position = "dodge", na.rm = TRUE, show.legend = FALSE) +
  labs(title = "Distribution of temperature by treatment",
       x = "Temperature",
       y = "Count")

ggplot(temperature, aes(x = temperature, fill = treatment)) +
  geom_density(alpha = 0.5, na.rm = TRUE, show.legend = FALSE) +
  labs(title = "Distribution of temperature by treatment",
       x = "Temperature")

# Q-Q plots
ggplot(temperature, aes(sample = temperature, color = treatment)) +
  stat_qq(show.legend = FALSE) +
  geom_abline(aes(intercept = mean(x = temperature), 
                  slope = sd(x = temperature))) +
  labs(title = "Q-Q plot of temperature by treatment",
   x = "Theoretical quantiles",
   y = "Sample quantiles",
   caption = "y-intercept is mean temperature, slope is standard deviation")

t.test(formula = temperature$temperature ~ temperature$treatment)
```

### E3 Temperature Summary:
Based on our graph comparing the means of the temperatures by treatment condition, we can see that the means for each condition are roughly the same. Based on our histograms and Q-Q plots, we can see that both treatment conditions are roughly normally distributed.
We are able to run independent t-tests here because our data consists of continuous variables, our observations are independent along with each group being independent, and the data are roughly normally distributed.
Based on our independent t-tests, there is not a statistically significant difference between vigilance rates in the control and noise treatments for the GT bird species [t(278)) = 0.67, p = 0.49]. The mean vigilance rate is likely to vary by -0.49 to 1.0 (95% CI).


```{r}
## HOURS FROM SUNRISE
sunrise_hours <- vigilance %>%
  select(site, species, treatment, hours.from.sunrise) 

sun_hours_summary <- sunrise_hours %>%
  group_by(treatment) %>%
  summarize(mean_sun_hours = mean(hours.from.sunrise),
            sd_sun_hours = sd(hours.from.sunrise),
            median_sun_hours = median(hours.from.sunrise),
            iqr = IQR(hours.from.sunrise),
            var = var(hours.from.sunrise),
            se = sd_sun_hours/sqrt(n()),
            CI_low = mean_sun_hours - 1.96*se,
            CI_high = mean_sun_hours + 1.96*se
  )
sun_hours_summary

summarized_sun_hours <- sun_hours_summary %>% 
  pivot_longer(cols = "mean_sun_hours":"CI_high",
               names_to = "stat",
               values_to = "values") %>%
  pivot_wider(names_from = treatment, values_from = values)

pretty_summary_sun <- kable(summarized_sun_hours, caption = "Summary statistics by treatment for hours from sunrise")
pretty_summary_sun


# Figure comparing the means
sun_hours_summary %>% 
  ggplot(aes(x = treatment, y = mean_sun_hours, fill = treatment)) +
  geom_col(show.legend = FALSE) +
  geom_errorbar(aes(ymin = CI_low, ymax = CI_high), width = 0.3, color = "black") +
  labs(title = "Mean hours from sunrise by noise treatment condition",
       x = "Treatment condition",
       y = "Mean hours from sunrise",
       caption = "error bars represent a 95% confidence interval")

# Histograms
ggplot(sunrise_hours, aes(x = hours.from.sunrise, fill = treatment)) +
  geom_histogram(binwidth = 0.5, position = "dodge", na.rm = TRUE, show.legend = FALSE) +
  labs(title = "Distribution of hours from sunrise by treatment",
       x = "Hours from sunrise",
       y = "Count")

ggplot(sunrise_hours, aes(x = hours.from.sunrise, fill = treatment)) +
  geom_density(alpha = 0.5, na.rm = TRUE, show.legend = FALSE) +
  labs(title = "Distribution of hours from sunrise by treatment",
       x = "Hours from sunrise")

# Q-Q plots
ggplot(sunrise_hours, aes(sample = hours.from.sunrise, color = treatment)) +
  stat_qq(show.legend = FALSE) +
  geom_abline(aes(intercept = mean(x = hours.from.sunrise), 
                  slope = sd(x = hours.from.sunrise))) +
  labs(title = "Q-Q plot of hours from sunrise by treatment",
   x = "Theoretical quantiles",
   y = "Sample quantiles",
   caption = "y-intercept is mean hours from sunrise, slope is standard deviation")

t.test(formula = sunrise_hours$hours.from.sunrise ~ sunrise_hours$treatment)
```

### E3 Hours fromn Sunrise Summary:
Based on our graph comparing the means of the hours after sunrise by treatment condition, we can see that the means for each condition are roughly the same. Based on our histograms and Q-Q plots, we can see that the control condition has two peaks, while the urban noise condition is roughly normally distributed.
Although the shape of the data isn't perfectly normal for the control condition, we ran independent t-tests here because our data consists of continuous variables, our observations are independent along with each group being independent.
Based on our independent t-tests, we can see that there is not a statistically significant difference between vigilance rates in the control and noise treatments for the GT bird species [t(287)) = 0.79, p = 0.42]. The mean vigilance rate is likely to vary by -0.2 to 0.6 (95% CI).
