---
title: "Homework Exercises Week 09"
author: "Alison Lawyer"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include = TRUE}
# keep this chunk in all your RMarkdown scripts
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 60), tidy = TRUE)
```

```{r}
# List required packages
library(tidyverse)
library(openxlsx)
library(car)
```



# HOMEWORK EXERCISES


## Exercise 3

In this exercise, you will use the dataset "visit_and_peck_rates.xlsx" from the same study to conduct ANOVAs for both visitation and peck rates, verify assumptions, visualize your data, and interpret your results. 


### Part 1

Import the dataset.
Create summary tables of visit and peck rates (you can create one table each or both together) with mean, median, standard error, 95% confidence interval, and sample size. 

```{r}
visit_peck <- read.xlsx("visit_and_peck_rates.xlsx")
colnames(visit_peck) <- sub(pattern = "\\.\\((.*?)\\)", replacement = '', colnames(visit_peck))

summary <- visit_peck %>%
  group_by(species, treatment) %>%
  summarize(mean_vr = mean(visit.rate),
            sd_vr = sd(visit.rate),
            median_vigilance = median(visit.rate),
            se_vr = sd_vr/sqrt(n()),
            CI_low_vr = mean_vr - 1.96*se_vr,
            CI_high_vr = mean_vr + 1.96*se_vr,
            mean_pr = mean(peck.rate),
            sd_pr = sd(peck.rate),
            median_pr = median(peck.rate),
            se_pr = sd_pr/sqrt(n()),
            CI_low_pr = mean_pr - 1.96*se_pr,
            CI_high_pr = mean_pr + 1.96*se_pr
  )
summary
```


### Part 2

Visualize your data. For each variable, create a plot showing the mean +- 95% CI by species. 
Be sure to create appropriate labels, legends, and error bars. 

```{r}
summary  %>%
  ggplot(aes(x = species, y = mean_vr, fill = treatment)) +
  geom_col(position = position_dodge(width = 0.9)) +
  geom_errorbar(aes(ymin = CI_low_vr, ymax = CI_high_vr), 
                width = 0.3, color = "black", 
                position = position_dodge(width = 0.9)) + 
  labs(title = "Mean visit rates for each species and noise condition",
       x = "Species",
       y = "Mean visit rates") +
  theme_minimal()

summary  %>%
  ggplot(aes(x = species, y = mean_pr, fill = treatment)) +
  geom_col(position = position_dodge(width = 0.9)) +
  geom_errorbar(aes(ymin = CI_low_pr, ymax = CI_high_pr), 
                width = 0.3, color = "black", 
                position = position_dodge(width = 0.9)) + 
  labs(title = "Mean peck rates for each species and noise condition",
       x = "Species",
       y = "Mean peck rates") +
  theme_minimal()
```


### Part 3

Conduct a two-way ANOVA for each variable to assess variation in mean visit and peck rates by species and condition. 
Print your summary ANOVA output.
Interpret the output. 

```{r}
two_way_vr <- aov(visit.rate ~ species * treatment, data = visit_peck)
two_way_pr <- aov(peck.rate ~ species * treatment, data = visit_peck)

summary(two_way_vr)
## The mean visit rate is very significantly different across species (F5,358 = 17.6, p < 0.05), but 
## not significantly different across treatment types (F1,358 = 3.4, p = 0.06), and
## not significantly different when you consider species and treatment types together 
## (F5,358 = 0.15, p = 0.97). 

summary(two_way_pr)
## The mean peck rate is very significantly different across species (F5,358 = 8.3, p < 0.05) and
## fairly significantly different across treatment types(F1,358 = 5.8, p = 0.01), 
## but not significant when you consider species and treatment types together (F5,358 = 1.1, p = 0.36). 
```


### Part 4

Verify that assumptions are met for ANOVA for both variables. 
Use appropriate techniques to assess normal distribution and homogeneity of variances. Describe what you see. 

```{r}
## First assumption: independent sampling of data. We can assume this is met. 

## Second assumption: normal distribution of residuals
plot(two_way_vr, 2)
plot(two_way_pr, 2)
## The residuals for each test are distributed roughly normally, except for the outliers on the far right end
## of the graphs.

## Third assumption: equal variance of residuals
leveneTest(visit.rate ~ species * treatment, data = visit_peck)
plot(two_way_vr, 3)
plot(two_way_vr, 1)
leveneTest(peck.rate ~ species * treatment, data = visit_peck)
plot(two_way_pr, 3)
plot(two_way_pr, 1)
## The third assumption is not met for both tests. The graphs and results from Levene's test demonstrate
## that there is a significant difference in the variances. 
```


### Part 5

Conduct post-hoc analyses if needed to identify the location of any significant differences. 
Summarize the result of your post-hoc analyses. 

```{r}
TukeyHSD(two_way_vr)
## The species with the largest significant differences in visit rate are:
## CT-CH 
## GT-CH
## R.-CT
## R.-GT
## There isn't a high significant difference between treatment types. 
## Comparing visit rate by species and treatment types, the most significant differences lie in:
## R.:control-GT:control
## R.:urban noise-GT:control
## R.:urban noise-GT:urban noise

TukeyHSD(two_way_pr)
## The species with the largest significant differences in visit rate are:
## R.-BT 
## R.-GT
## R.-NH
## There is a higher significant difference between treatment types for peck rate than visit rate.
## Comparing visit rate by species and treatment types, the most significant differences lie in:
## R.:urban noise-NH:control
## R.:urban noise-BT:control
## R.:urban noise-NH:control

## These results suggest that treatment type has a higher significance for peck rate than visit rate, 
## while species is the main influencing factor on visit rate. 
```


